A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-08-27 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewing conclusion

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts \cite{plavala2023,ariano2017,caticha2019,hall2016}, [CITE - GPT, Ariel, Reginatto/Hall...] which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics\cite{chsh1969}.[CITE] Quasiprobabilistic approaches \cite{ferrie2010,mandel1995,monchietti2025} [CITE] are also used in quantum mechanics, but it is not completely clear how exactly they relate to the standard one.

@@

We will then review some results from the use of quasiprobability measures. The most used is the Wigner function,[cite] that represents each state as a signed probability measure over phase space, and expectations of observables are recovered through standard integration. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand\cite{cahill1969}. [cite] The key insight is that it is not negative probability that characterizes quasiprobability,\footnote{Some representations guarantee the quasiprobability to be non-negative.} but the lack of mutual exclusivity of the points of the space. The quasiprobability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

@@

Lastly, we lay down the requirements for a non-additive generalization of probability theory, which should be in principle possible. Since there is a wide literature on the subject, [cite Grabish, Pap, TorraSugeno] we only show that the most used non-additive integrals, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

@@

We have seen how we can give a unified picture for how different measure theoretic tools can be used to represent states in physical theories. Standard probability theory can be understood as representing ensembles as mixtures of pure states, as extreme points. In particular, classical probability applies under the following equivalent conditions:
\begin{enumerate}
	\item the space of ensembles is a Choquet simplex
	\item every probability distribution over the pure states identifies a unique ensemble
	\item every ensembles is uniquely decomposable into pure states
	\item sets of pure states identify mutually exclusive events
	\item the fraction capacity of every ensemble is additive.
\end{enumerate}
Quantum mechanics fails to be characterized by classical probability precisely because it does not satisfies any of these conditions. We also saw that it is the embedding of any space of ensembles into a vector space that allows the use of quasi-probability, which is mandated by very minimal physical assumptions. However, quasi-probability representations are not unique, and therefore are not suitable as foundational descriptions.

We have seen, instead, that non-additive measures can be created to generalize probability and state counting measure to the general case. Intuitively, the fraction capacity simply tells us how much of a particular ensemble can be understood as coming from a mixture of a set of other ensembles. The state capacity, instead, is highest exponential entropy achievable by mixing a given sets of ensembles, which reduces to the Liouville measure of classical mechanics and the dimensionality for quantum subspaces.

>>>>

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts \cite{plavala2023,ariano2017,caticha2019,hall2016}, [CITE - GPT, Ariel, Reginatto/Hall...] which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.[CITE] Quasiprobabilistic approaches [CITE] are also used in quantum mechanics, but it is not completely clear how exactly they relate to the standard one.

@@

We will then review some results from the use of quasiprobability measures. The most used is the Wigner function,[cite] that represents each state as a signed probability measure over phase space, and expectations of observables are recovered through standard integration. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand.[cite] The key insight is that it is not negative probability that characterizes quasiprobability,\footnote{Some representations guarantee the quasiprobability to be non-negative.} but the lack of mutual exclusivity of the points of the space. The quasiprobability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

@@

Lastly, we lay down the requirements for a non-additive generalization of probability theory, which should be in principle possible. Since there is a wide literature on the subject, [cite] we only show that the most used non-additive integrals, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

@@

We have seen how we can give a unified picture for how different measure theoretic tools can be used to represent states in physical theories. Standard probability theory can be understood as representing ensembles as mixtures of pure states, as extreme points. In particular, classical probability applies under the following equivalent conditions:
\begin{enumerate}
	\item the space of ensembles is a Choquet simplex
	\item every probability distribution over the pure states identifies a unique ensemble
	\item every ensemble is uniquely decomposable into pure states
	\item sets of pure states identify mutually exclusive events
	\item the fraction capacity of every ensemble is additive.
\end{enumerate}
Quantum mechanics fails to be characterized by classical probability precisely because it does not satisfy any of these conditions. We also saw that it is the embedding of any space of ensembles into a vector space that allows the use of quasi-probability, which is mandated by very minimal physical assumptions. However, quasi-probability representations are not unique, and therefore are not suitable as foundational descriptions.

We have seen, instead, that non-additive measures can be created to generalize probability and state counting measure. Intuitively, the fraction capacity simply tells us how much of a particular ensemble can be understood as coming from a mixture of a set of other ensembles. The log of the state capacity, instead, is the highest entropy achievable by mixing a given set of ensembles, which reduces to the Liouville measure of classical mechanics and the dimensionality for quantum subspaces.
